---
title: "Applied Data Analysis Homework 4"
author: "Kevin Sanchez"
date: "5/12/2020"
output: html_document
---

``` {r}
library(dplyr)
library(tidyverse)
library(patchwork)
library(scales)
library(collapse)
library(BBmisc)
library(devtools)
remotes::install_github("joachim-gassen/tidycovid19")
library(tidycovid19)

merged <- download_merged_data(cached = TRUE)
merged <- merged %>%
  group_by(country, iso3c) %>%
  arrange(country, iso3c, date) %>%
  mutate(
    confirmed = ifelse(is.na(confirmed), 0, confirmed),
    deaths = ifelse(is.na(deaths), 0, deaths),
    recovered = ifelse(is.na(recovered), 0, recovered)
  ) %>%
  mutate(
    daily_confirmed = confirmed - lag(confirmed, n = 1),
    daily_deaths = deaths - lag(deaths, n = 1),
    daily_recovered = recovered - lag(recovered, n = 1)
  ) %>%
  mutate(
    daily_confirmed = replace_na(daily_confirmed, 0),
    daily_deaths = replace_na(daily_deaths, 0),
    daily_recovered = replace_na(daily_recovered, 0)
  ) %>%
  ungroup() %>%
  arrange(country, iso3c, date)

add_world1 <- merged %>%
  group_by(date) %>%
  arrange(date) %>%
  summarize(
    country = "World", iso3c = NA,
    confirmed = sum(confirmed, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    recovered = sum(recovered, na.rm = TRUE),
    timestamp = fmode(timestamp)
  ) %>%
  mutate(
    daily_confirmed = confirmed - lag(confirmed, n = 1),
    daily_deaths = deaths - lag(deaths, n = 1),
    daily_recovered = recovered - lag(recovered, n = 1)
  ) %>%
  mutate(
    daily_confirmed = replace_na(daily_confirmed, 0),
    daily_deaths = replace_na(daily_deaths, 0),
    daily_recovered = replace_na(daily_recovered, 0)
  ) %>%
  ungroup() %>%
  arrange(country, iso3c, date)

add_world2 <- merged %>%
  group_by(country, iso3c) %>%
  summarize(
    population = fmode(population),
    land_area_skm = fmode(land_area_skm),
    timestamp = fmode(timestamp)
  ) %>%
  ungroup() %>%
  summarize(
    country = "World", iso3c = NA,
    population = sum(population, na.rm = TRUE),
    land_area_skm = sum(land_area_skm, na.rm = TRUE)
  ) %>%
  mutate(pop_density = population / land_area_skm)

add_world <- left_join(add_world1, add_world2, by = c("country", "iso3c"))
merged <- bind_rows(merged, add_world)

cv_data <- pivot_longer(merged,
  cols = c(
    "confirmed", "deaths", "recovered",
    "daily_confirmed", "daily_deaths", "daily_recovered"
  ),
  names_to = "variable", values_to = "cases"
) %>%
  arrange(country, variable, date) %>%
  rename(area = land_area_skm, density = pop_density) %>%
  mutate(rate = cases / population * 10^6) %>%
  filter(date < "2020-05-11")

cv_summary <- function(d, country_list = "World",
                       plot = TRUE, facet = "country",
                       status = c("confirmed", "deaths", "recovered")) {
  
  my_palette <- c(
    "#5B1A18", "#FD6467", "#F1BB7B", "#D67236",
    "#0F0D0E", "#9986A5", "#79402E", "#CCBA72", "#D9D0D3", "#8D8680",
    "#EAD3BF", "#AA9486", "#B6854D", "#39312F", "#1C1718"
  )

  if (facet == "country") {
    fill <- "variable"
    n <- length(unique(d$variable)) / 2
  }

  if (facet == "variable") {
    fill <- "country"
    n <- length(country_list)
  }

  if ("All" %in% country_list) {
    country_list <- unique(d$country)
    country_list <- setdiff(country_list, "World")
  }

  if ("World" %in% country_list) {
    d <- d %>% filter(country %in% country_list)

    totals <- d %>%
      group_by(variable) %>%
      summarize(
        country = "World",
        cases = max(cases),
        population = max(population),
        area = max(area),
        density = max(density),
        rate = max(rate, na.rm = TRUE),
        on = max(date)
      ) %>%
      select(country, variable, cases, population, area, density, rate, on) %>%
      arrange(variable) %>%
      ungroup()
  }

  if ("World" %nin% country_list) {
    d <- d %>% filter(country %in% country_list)
    totals <- d %>%
      group_by(country, variable) %>%
      summarize(
        cases = max(cases),
        population = max(population),
        area = max(area),
        density = max(density),
        rate = max(rate, na.rm = TRUE),
        on = max(date),
        gdp_capita = fmode(gdp_capita),
        income = fmode(income),
        life_expectancy = fmode(life_expectancy),
        max_sd = max(soc_dist),
        max_mr = max(mov_rest)
      ) %>%
      select(
        country, variable, cases, population, area, density, rate,
        gdp_capita, income, life_expectancy, max_sd, max_mr, on
      ) %>%
      arrange(country, variable) %>%
      ungroup()
  }

  if (plot == TRUE) {
    cc <- filter(d, variable %in% status)
    cum_cases_plot <- ggplot(
      data = cc,
      aes(
        x = date, y = cases + 1, color = .data[[fill]],
        fill = .data[[fill]]
      )
    ) +
      geom_point(size = 0.5) +
      geom_line() +
      facet_wrap(~ .data[[facet]], ncol = 5) +
      xlab("Date") +
      ylab("Log Cumulative Cases") +
      scale_y_log10(
        breaks = trans_breaks("log10", function(x) 10^x),
        labels = trans_format("log10", math_format(10^.x))
      ) +
      scale_color_manual(
        aesthetics = c("color", "fill"),
        name = NULL, values = my_palette
      )

    dc <- filter(d, variable %in% paste0("daily_", status))
    daily_cases_plot <- ggplot(
      data = dc,
      aes(
        x = date, y = cases, color = .data[[fill]],
        fill = .data[[fill]]
      )
    ) +
      geom_point(size = 0.5) +
      geom_line() +
      facet_wrap(~ .data[[facet]], ncol = 5) +
      xlab("Date") +
      ylab("Daily Cases") +
      scale_color_manual(
        aesthetics = c("color", "fill"),
        name = NULL, values = my_palette
      )
  }

  if (plot == TRUE) {
    return(list(
      totals = totals,
      cum_cases_plot = cum_cases_plot,
      daily_cases_plot = daily_cases_plot
    ))
  } else {
    return(list(totals = totals))
  }
}
```

## Challenge 1
``` {r}
cv_summary(cv_data)
```

## Challenge 2
``` {r}
cv_summary(cv_data, country_list = c("US", "United Kingdom", "Canada", "France",
  "Germany", "Italy", "Japan", "China", "Russia", "Iran"), facet = "country")
cv_summary(cv_data, country_list = c("US", "United Kingdom", "Canada", "France",
  "Germany", "Italy", "Japan", "China", "Russia", "Iran"), facet = "variable")
```

## Challenge 3
``` {r}
d <- cv_summary(cv_data, country_list = "All", plot = FALSE)
d <- d[[1]]
d <- filter(d, population > 1000000)
length(unique(d$country))
```
There are 153 countries that have a population over 1,000,000 with confirmed coronavirus 
cases (Puerto Rico is not a country).

## Challenge 4
``` {r}
overall <- filter(d, variable == "confirmed")
overall <- arrange(overall, desc(rate))
head(overall, 10)
overall <- arrange(overall, desc(cases))
head(overall, 10)

daily <- filter(d, variable == "daily_confirmed")
daily <- arrange(daily, desc(rate))
head(daily, 10)
daily <- arrange(daily, desc(cases))
head(daily, 10)
```

## Challenge 5
For this challenge, I will filter the dataset **overall** to include just the variables of
interest. 5 countries were removed from the dataset due to either NAs or -Inf values. This
new dataset is **overall_filt_ex**.
``` {r}
overall <- overall[-154,]
overall_filt <- select(overall, country, population, density, rate, gdp_capita, income)
overall_filt_ex <- na.exclude(overall_filt)
overall_filt_ex <- overall_filt_ex[-149, ]

summary(lm(rate ~ density + population + gdp_capita + income, data = overall_filt_ex))
```
Now, I will look at each set of variables individually to determine if a logarithmic
transformation is necessary. 
```{r}
summary(lm(rate ~ density, overall_filt_ex))
RvD <- ggplot(overall_filt_ex, aes(x = density, y = rate)) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  xlab("Density") + ylab("Rate")
RvD

log_RvD <- ggplot(overall_filt_ex, aes(x = log(density), y = log(rate))) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  xlab(" log of Density") + ylab("log of Rate")
log_RvD

summary(lm(log(rate) ~ log(density), overall_filt_ex))

summary(lm(rate ~ population, overall_filt_ex))
RvP <- ggplot(overall_filt_ex, aes(x = population, y = rate)) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  xlab("Population") + ylab("Rate")
RvP

log_RvP <- ggplot(overall_filt_ex, aes(x = log(population), y = log(rate))) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  xlab(" log of Population") + ylab("log of Rate")
log_RvP

summary(lm(log(rate) ~ log(population), overall_filt_ex))

summary(lm(rate ~ gdp_capita, overall_filt_ex))
RvG <- ggplot(overall_filt_ex, aes(x = gdp_capita, y = rate)) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  xlab("GDP per Capita") + ylab("Rate")
RvG

log_RvG <- ggplot(overall_filt_ex, aes(x = log(gdp_capita), y = log(rate))) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  xlab(" log of GDP per Capita") + ylab("log of Rate")
log_RvG

summary(lm(log(rate) ~ log(gdp_capita), overall_filt_ex))

overall_filt_ex$income <- factor(overall_filt_ex$income,
    levels = c('Low income','Lower middle income','Upper middle income','High income'),
    ordered = TRUE)

summary(lm(rate ~ income, overall_filt_ex))
RvI <- ggplot(overall_filt_ex, aes(x = income, y = rate)) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE, aes(group = 1), color = "black") +
  xlab("Income") + ylab("Rate")
RvI

log_RvI <- ggplot(overall_filt_ex, aes(x = income, y = log(rate))) +
  geom_point() + 
  geom_smooth(method="lm", se = FALSE, aes(group = 1), color = "black") + 
  xlab("Income") + ylab("log of Rate")
log_RvI

summary(lm(log(rate) ~ income, overall_filt_ex))
```
Based on these graphs and linear regression analyses, it appears that both income and GDP
per capita are significantly correlated with the rate of coronavirus transmission.

## Challenge 6
``` {r}
library(MASS)
overall_lm_opt <- lm(rate ~ log(density) + log(population) + log(gdp_capita) + income, 
             data = overall_filt_ex)
overall_AIC <- stepAIC(overall_lm_opt, scope = . ~ ., direction = "both")

library(MuMIn)
r.squaredGLMM(overall_lm_opt)
r.squaredGLMM(overall_AIC)
```
The best model based on the `stepAIC()` function removes the **log(population)** variable. 
The pseduo-R<sup>2</sup> value slightly increased from 0.4745 to 0.4763.
``` {r}
daily_ex <- na.exclude(daily)
daily_lm_opt <- lm(rate ~ log(density) + log(population) + log(gdp_capita) + income, 
             data = daily_ex)
daily_AIC <- stepAIC(daily_lm_opt, scope = . ~ ., direction = "both")

r.squaredGLMM(daily_lm_opt)
r.squaredGLMM(daily_AIC)
```
Repeating this model using the **daily_ex** dataset revealed that the best model does not 
include the **income** and **log(population)** variables. Therefore, this model is 
slightly different than when using the **overall_filt_ex** dataset. The 
pseudo-R<sup>2</sup> value actually decreased from 0.2501 to 0.2334, which was unexpected. 

## Challenge 7
``` {r}
daily_lm_max <- lm(data = daily_ex, rate ~ log(density) + log(gdp_capita) + max_sd + max_mr)
daily_AIC_max <- stepAIC(daily_lm_max, scope = . ~ ., direction = "both")

r.squaredGLMM(daily_lm_max)
r.squaredGLMM(daily_lm_opt)
```
From this analysis with the variables **max_sd** and **max_mr** added, it was revealed 
that only **max_mr** and not **max_sd** helped create a better model. The 
pseudo-R<sup>2</sup> value minimally increased from 0.2470 to 0.2501. 

## Challenge 8
``` {r}
cv_data_filt <- filter(cv_data, 
                       population > 1000000, 
                       rate > 1, 
                       variable == "daily_confirmed")

library(lme4)
m1 <- lmer(data = cv_data_filt, rate ~ log(density) + log(gdp_capita) + soc_dist + 
          mov_rest + (1 | country) + (1 | date))

lme0 <- lmer(data = cv_data_filt, rate ~ (1 | country) + (1 | date), REML = FALSE)
lme1 <- lmer(data = cv_data_filt, rate ~ log(density) + (1 | country) + (1 | date), 
             REML = FALSE)
lme2 <- lmer(data = cv_data_filt, rate ~ log(gdp_capita) + (1 | country) + (1 | date), 
             REML = FALSE)
lme3 <- lmer(data = cv_data_filt, rate ~ soc_dist + (1 | country) + (1 | date), 
             REML = FALSE)
lme4 <- lmer(data = cv_data_filt, rate ~ mov_rest + (1 | country) + (1 | date), 
             REML = FALSE)
lme5 <- lmer(data = cv_data_filt, rate ~ log(density) + log(gdp_capita) + soc_dist + 
          mov_rest + (1 | country) + (1 | date), REML = FALSE)
lme6 <- lmer(data = cv_data_filt, rate ~ log(density) + log(gdp_capita) + soc_dist + 
            (1 | country) + (1 | date), REML = FALSE)
lme7 <- lmer(data = cv_data_filt, rate ~ log(density) + log(gdp_capita) + mov_rest + 
            (1 | country) + (1 | date), REML = FALSE)
lme8 <- lmer(data = cv_data_filt, rate ~ log(density) + soc_dist + mov_rest + 
            (1 | country) + (1 | date), REML = FALSE)
lme9 <- lmer(data = cv_data_filt, rate ~ log(gdp_capita) + soc_dist + mov_rest + 
            (1 | country) + (1 | date), REML = FALSE)
lme10 <- lmer(data = cv_data_filt, rate ~ log(density) + log(gdp_capita) + (1 | country) +
            (1 | date), REML = FALSE)
lme11 <- lmer(data = cv_data_filt, rate ~ log(density) + soc_dist + (1 | country) + 
            (1 | date), REML = FALSE)
lme12 <- lmer(data = cv_data_filt, rate ~ log(density) + mov_rest + (1 | country) + 
            (1 | date), REML = FALSE)
lme13 <- lmer(data = cv_data_filt, rate ~ log(gdp_capita) + soc_dist + (1 | country) + 
            (1 | date), REML = FALSE)
lme14 <- lmer(data = cv_data_filt, rate ~ log(gdp_capita) + mov_rest + (1 | country) + 
            (1 | date), REML = FALSE)
lme15 <- lmer(data = cv_data_filt, rate ~ soc_dist + mov_rest + (1 | country) + 
            (1 | date), REML = FALSE)

library(AICcmodavg)
lme_aic <- aictab(list(lme0, lme1, lme2, lme3, lme4, lme5, lme6, lme7, lme8, lme9, lme10,
                       lme11, lme12, lme13, lme14, lme15))
lme_aic
```
The best model from this analysis is **lme6** which only excludes the variable 
**mov_rest** from the model.  
```{r}
r.squaredGLMM(lme6)
r.squaredGLMM(lme1)
```
The pseudo-R<sup>2</sup> value of **lme6** was 0.5366, which was higher than the 
pseudo-R<sup>2</sup> value of 0.4686 for **lme1** (the model with the highest AIC value). 